Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-12 fscd19/fscd19.tex
[ fscd19 ] reflection for polarized lambda

Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ shall be constructed by the
following rules:
\begin{gather*}
  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

@@

Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
contain negative normal forms.

Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
starting from a forced $\tThunk$ rather than from variables of negative
types (as those do not exist in CBPV).  However, due to normality
the $\tThunk$ cannot be a $\tthunk$, but only a variable
$\Thunk N \in \Gamma$.
\begin{gather*}
  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Variables are originally introduced by either $\tabs$ or
the $\tbind$ind of a neutral of type $\Comp P$ to a new variable of type
$P$.  Variables of composite value type can be broken down by pattern
matching, introducing variables of smaller type.  These positive
eliminations plus $\tbind$ are organized in the
inductively defined strong monad \fbox{$\Cov$}.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\qquad
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
$\Comp P$ is reached.  Then, elimination of neutrals and variables is
possible through the $\Cov$ monad until an answer can be given in form
of a base neutral ($\tNe\,\natom$) or a normal value.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tret}
      {\CoverP {\tVal[\tNf]\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\nunit}{}{\Nf \top \Gamma}
\qquad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
negative semantic values
$\trun : \Cov \den N \todot \den N$.

Reification $\down P : \den P \todot \tVal[\tNf]\,P$ at positive types produces
a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.  Luckily, as $\down N$ produces a normal
form, monadic intermediate computations are permitted under a final
$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
%
In the following definition of reflection we use the abbreviation
\fbox{\(
  \tfresh^P_\Gamma
    = \up[\cext \Gamma P] P \vv0
    : \Cover{\den P}{(\cext \Gamma P)}
\)}.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
  \upG {\patom}         & x & = & x \\
  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
    \left(
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
     \star
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    \right) \\
    % (\tzip
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    % ) \\
  \upG{\tyzero}         & x & = & \tabort\;x \\
  \upG{P_1 + P_2}       & x & = & \tcase\;x
    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
    \\
  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
\end{array}
\,
\begin{array}{lcl}
  \down P & : & \den P \todot \tVal[\tNf]\,P \\
  \down {\patom} & = & \tvar \\
  \downG {\tyone} () & = & \punit \\
  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
    \,(\downG {P_1} a_1)
    \,(\downG {P_2} a_2)
    \\
  \down {\tyzero} & = & \tmagic \\
  \down {P_1 + P_2} & = &
    [  \tinj_1 \comp \down{P_1}
    ,\ \tinj_2 \comp \down{P_2}
    ] \\
  \down {\Thunk N} & = & \tthunk \circ \down N \\
\end{array}
\]
Reflection at positive pairs uses monoidal functoriality
$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
which for monads $\C$ can be defined by
$c_1 \star c_2 =
   \tjoin \; (\tmap\;(\lambda\,a_1.\
   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
    \;c_2)
    \;c_1)$.
% $c_1 \star c_2 =
%    \tjoin \; (\tmap\;(\lambda\,a_1.\
%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
%     \;c_2))
%     \;c_1)$.
% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.

For negative product types, reflection and reification works as before:
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
  \upG \natom & u & = & \treturn\,u \\
  \upG \top   & u & = & \ttempty \\
  \upG {N_1 \ntimes N_2} & u & = & \left(
    \upG {N_1} \; (\prj 1 u)
    ,\
    \upG {N_2} \; (\prj 2 u)
    \right) \\
\end{array}
\
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
  \downG \natom & c  & = & \tne\,c \\
  \downG \top   & () & = & \nunit \\
  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
    \, (\downG {N_1} \; b_1)
    \, (\downG {N_2} \; b_1)
    \\
\end{array}
\]
Reflection for function types is also unchanged, except that $\tapp$
expects a value argument now.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \upG {P \To N} & u & = & \lambda
    \;(\tau : \Gamma \sublist \Delta)
    \;(a : \den P \Delta)
    . \
    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
    \\
  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
    \stmap
      \left(
        \lambda
          \;(\tau : (\cext \Gamma P) \sublist \Delta)
          \;(a : \den P \Delta)
          . \
          f\; (\twk^P \rcomp \tau)\; a
        \right)
      \,\tfresh^P_\Gamma
    \right) \right)
    \\
\end{array}
\]
The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
can only be generated in the monad, due to monadic positive
reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \treturn\,() \\
  \idenv{\cext \Gamma P} & = &
    (\tren\,\twk^P\,\idenv\Gamma) \star \tfresh^P_\Gamma \\
\end{array}
\]
Putting things together, we obtain the normalization function
\[
  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
    \left(
      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
    \right)
.
\]

Taking stock, we have arrived at normal forms that eagerly introduce
($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
introduce positive types ($\tVal$).  However, the elimination of
positive types is still rather non-deterministic.
It is possible to only partially break up a composite positive type
and leave smaller, but still composite positive types for later
pattern matching.

The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
constrained by the presence of the split variable

@@

Polarized lambda calculus \cite{espiritoSanto:entcs17}
divides types deeply into positive and negative.

\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    \mid \tyzero \mid P_1 + P_2
    \mid \tyone \mid P_1 \times P_2
    \mid \Thunk N
  & \mbox{positive type}
\\
  \NTy & \ni & N,M & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
  & \mbox{negative type}
\\
  \HTy & \ni & H & ::= & \patom \mid P
  & \mbox{hypothesis type}
\\
\end{array}
\]
Espirito Santo's version actually does not contain the positive
product $\times$, however, it appears in Zeilberger
\cite{zeilberger:PhD}.

Polarized lambda-calculus is a focused calculus, it eagerly employs
invertible rules: the introduction rules for negative types and the
elimination rules for positive types.  As a consequence of the latter,
variables are either of atomic or negative type ($\HTy$).
Contexts $\Gamma,\Delta$ are lists of $\HTy$s.
%\fbox{$X ::= \patom \mid P$}.

@@

@@

Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  It relies on the previous definitions
with instance $\T = \tNf$.
\begin{gather*}
  \nru{\tne}
      {\Cov[\tNf](\tNe[\tNf]\,\natom)\,\Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tval}
      {\Cov[\tNf](\tVal[\tNf]\,P)\,\Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\end{gather*}

>>>>

Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ shall be constructed by the
following rules:
\begin{gather*}
  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

@@

Positive normal forms are values $v : \VNf P \Gamma$
referring only to atomic variables and
whose thunks only
contain negative normal forms.
% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
% contain negative normal forms.
\begin{gather*}
  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
\qquad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
      {\PValNfP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
\end{gather*}

Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
starting from a forced $\tThunk$ rather than from variables of negative
types (as those do not exist in CBPV).  However, due to normality
the $\tThunk$ cannot be a $\tthunk$, but only a variable
$\Thunk N \in \Gamma$.
\begin{gather*}
  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Variables are originally introduced by either $\tabs$ or
the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
$P$.  Variables of composite value type can be broken down by pattern
matching, introducing variables of smaller type.  These positive
eliminations plus $\tbind$ are organized in the
inductively defined strong monad \fbox{$\Cov$}.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\qquad
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
$\Comp P$ is reached.  Then, elimination of neutrals and variables is
possible through the $\Cov$ monad until an answer can be given in form
of a base neutral ($\tNe\,\natom$) or a normal value.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\nunit}{}{\Nf \top \Gamma}
\qquad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
negative semantic values
$\trun : \Cov \den N \todot \den N$.

Reification $\down P : \den P \todot \tVNf\,P$ at positive types produces
a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.  Luckily, as $\down N$ produces a normal
form, monadic intermediate computations are permitted under a final
$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
%
In the following definition of reflection we use the abbreviation
\fbox{\(
  \tfresh^P_\Gamma
    = \up[\cext \Gamma P] P \vv0
    : \Cover{\den P}{(\cext \Gamma P)}
\)}.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
  \upG {\patom}         & x & = & x \\
  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
    \left(
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
     \star
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    \right) \\
    % (\tzip
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    % ) \\
  \upG{\tyzero}         & x & = & \tabort\;x \\
  \upG{P_1 + P_2}       & x & = & \tcase\;x
    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
    \\
  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
\end{array}
\,
\begin{array}{lcl}
  \down P & : & \den P \todot \tVNf\,P \\
  \down {\patom} & = & \tvar \\
  \downG {\tyone} () & = & \punit \\
  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
    \,(\downG {P_1} a_1)
    \,(\downG {P_2} a_2)
    \\
  \down {\tyzero} & = & \tmagic \\
  \down {P_1 + P_2} & = &
    [  \tinj_1 \comp \down{P_1}
    ,\ \tinj_2 \comp \down{P_2}
    ] \\
  \down {\Thunk N} & = & \tthunk \circ \down N \\
\end{array}
\]
Reflection at positive pairs uses monoidal functoriality
$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
which for monads $\C$ can be defined by
$c_1 \star c_2 =
   \tjoin \; (\tmap\;(\lambda\,a_1.\
   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
    \;c_2)
    \;c_1)$.
% $c_1 \star c_2 =
%    \tjoin \; (\tmap\;(\lambda\,a_1.\
%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
%     \;c_2))
%     \;c_1)$.
% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.

For negative product types, reflection and reification works as before:
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
  \upG {\Comp P} & u & = & \tbind\,u\,\tfresh^P_\Gamma \\
  \upG \natom    & u & = & \treturn\,u \\
  \upG \top      & u & = & \ttempty \\
  \upG {N_1 \ntimes N_2} & u & = & \left(
    \upG {N_1} \; (\prj 1 u)
    ,\
    \upG {N_2} \; (\prj 2 u)
    \right) \\
\end{array}
\
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
  \downG {\Comp P} & c  & = & \tmap\;(\down P)\;c \\
  \downG \natom    & c  & = & \tne\,c \\
  \downG \top      & () & = & \nunit \\
  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
    \, (\downG {N_1} \; b_1)
    \, (\downG {N_2} \; b_2)
    \\
\end{array}
\]
Reflection for function types is also unchanged, except that $\tapp$
expects a value argument now.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \upG {P \To N} & u & = & \lambda
    \;(\tau : \Gamma \sublist \Delta)
    \;(a : \den P \Delta)
    . \
    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
    \\
  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
    \stmap
      \left(
        \lambda
          \;(\tau : (\cext \Gamma P) \sublist \Delta)
          \;(a : \den P \Delta)
          . \
          f\; (\twk^P \rcomp \tau)\; a
        \right)
      \,\tfresh^P_\Gamma
    \right) \right)
    \\
\end{array}
\]
The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
can only be generated in the monad, due to monadic positive
reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \treturn\,() \\
  \idenv{\cext \Gamma P} & = &
    (\tren\,\twk^P\,\idenv\Gamma) \star \tfresh^P_\Gamma \\
\end{array}
\]
Putting things together, we obtain the normalization function
\[
  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
    \left(
      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
    \right)
.
\]

Taking stock, we have arrived at normal forms that eagerly introduce
($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
introduce positive types ($\tVnf$).  However, the elimination of
positive types is still rather non-deterministic.
It is possible to only partially break up a composite positive type
and leave smaller, but still composite positive types for later
pattern matching.
The last refinement, chaining also the positive eliminations, will be
discussed in the following section.
% The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
% constrained by the presence of the split variable

@@

% Polarized lambda calculus \cite{espiritoSanto:entcs17}
% divides types deeply into positive and negative.

% \[
% \begin{array}{lllrl@{\qquad}l}
%   \PTy & \ni & P,Q & ::= & \patom
%     \mid \tyzero \mid P_1 + P_2
%     \mid \tyone \mid P_1 \times P_2
%     \mid \Thunk N
%   & \mbox{positive type}
% \\
%   \NTy & \ni & N,M & ::= & \natom
%     \mid \top \mid N_1 \ntimes N_2
%     \mid P \To N
%     \mid \Comp P
%   & \mbox{negative type}
% \\
%   \HTy & \ni & H & ::= & \patom \mid N
%   & \mbox{hypothesis type}
% \\
% \end{array}
% \]
% Espirito Santo's version actually does not contain the positive
% product $\times$, however, it appears in Zeilberger
% \cite{zeilberger:PhD}.

Polarized lambda-calculus
\cite{zeilberger:PhD,espiritoSanto:entcs17}
is a focused calculus, it eagerly employs
invertible rules: the introduction rules for negative types and the
elimination rules for positive types.  As a consequence of the latter,
variables are either of atomic or negative type $H ::= \patom \mid N$.
%($\HTy$).
Contexts $\Gamma,\Delta$ are lists of $H$s.
%$\HTy$s.
%\fbox{$X ::= \patom \mid P$}.

@@

Focalization is a technique to remove don't-care non-determinism from
proof search, and as such, polarized lambda calculus is foremost a
calculus of normal forms.
These normal forms are given by four mutually defined inductive
families of presheaves $\tVnf\,P$,
$\tNe\,N$, $\Cov\,\J$,
and $\tNf\,N$.  As they are very similar to the CBPV normal forms
given in the last section we only report the differences.

Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.

Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
of with $\tforce$, as forcing thunks is already performed in $\nhyp$
when adding analysing hypotheses of $\tThunk$ type.
\[
  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
\]

The Cover monad \fbox{$\Cover \J \Gamma$}
lacks constructors $\tsplit$, $\tcase$ and $\tabort$
since the pattern matching is taken care of by $\Add P$.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
      {\Cover \J \Gamma}
\end{gather*}

The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
with the exception that pattern matching happens eagerly in $\tabs$.
\[
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\]

All these inductive families are presheaves, due to factored
presentation using $\Add P$ and $\Cov$ the proof is not a simple
mutual induction.  Yet, in Agda, the generic proof goes through using a sized
typing for these inductive families.  Similarily, defining the
$\tjoin$ for monad $\Cov$ relies on sized typing.
\[
\begin{array}{lll}
  \tjoin & : & \forall i.\
    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
  \tjoin[i+1]\,(\tbind^i\;t\;k)
    & = & \tbind^\infty\;
     (t : \Ne P \Gamma)\;
     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
       \right)
     \right)
\end{array}
\]
Herein, we used the sized typing of the constructors of $\Cov$:
\[
\begin{array}{lll}
  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
  \tbind  & : & \forall i.\
    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
\end{array}
\]

Due to the eager splitting of positive hypotheses, reflection at type $P$
now lives in the $\Add P$ graded monad rather than $\Cov$.  Further,
as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
produce one positive semantic value; instead, one such value is needed
for every branch.  We implement
\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
as a higher-order function expecting a continuation $k$ which is
invoked for each generated branch with the semantic value of type $P$
constructed for this branch.
\[
\begin{array}{l@{~}lcl}
  \reflect {\patom} & k & = &
    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
  \reflect {\Thunk N} & k & = &
    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
  \reflect {\tyzero} & k & = & \nbranch \\
  \reflect {P_1 + P_2} & k & = & \bbranch
    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
    \\
  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
  \reflect{P_1 \times P_2} & k & = & \bsplit
    \,\left(
      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
        \right)
      \right)
    \right)
\end{array}
\]
Reflecting at a positive atomic type $\patom$ is the regular ending of a
reflection pass: we call continuation $k$ with a fresh variable
$\pvar\,\tzero$ of type $\patom$, making space for the variable
using $\twk^{\patom}$.
In case we end at type $\Thunk N$, we add a new variable
$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
$\eta$-expansion via $\up N$.  Two more endings are possible:
At type
$\tyzero$, we have reached an absurd case, meaning that no
continuation is necessary since we can conclude with
\emph{ex falsum quod libet}.
At type $\tyone$, there is no need to add a new variable, as values
of type $\tyone$ contain no information.  We simply pass the unit
value $\ttempty$ to $k$ in this case.
Reflecting at $P_1 + P_2$ generates two branches, which may result in
several uses of the continuation $k$.
In the first branch, we recursively reflect at $P_1$.  Its
continuation will receive a semantic value in $\den{P_1}$ which we
inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
second branch proceeds analogously.
Finally reflecting at $P_1 \times P_2$ means we first have to analyze
$P_1$, and in each of the generated branches we continue to analyze
$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
$\reflect{P_1}$.  Each reflection phase gives us a semantic value
$a_i$ of type $P_i$, which we combine to a tuple before passing it to
$k$.  Note also that the context extension $\tau_1$ created in the
first phase needs to be composed with the context extension $\tau_2$
of the second phase to transport $k$ into the final context.  Further,
the value $a_1$ was constructed relative to the target of
$\tau_1$ and still needs
to be transported with $\tau_2$ before paired up with $a_2$.

\subsection{Old, explicit syntax}

@@

Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  It relies on the previous definitions
with instance $\T = \tNf$.
\begin{gather*}
  \nru{\tne}
      {\Cov[\tNf](\tNe[\tNf]\,\natom)\,\Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tval}
      {\Cov[\tNf](\tVNf\,P)\,\Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\end{gather*}
